---
title: "Sequencing Coverage QC Report"
subtitle: "Whole Genome Sequencing — Mosdepth Summary"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    smooth-scroll: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 5
params:
  config_file: "config/test_plot_config.yaml"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(plotly)
library(yaml)
library(DT)
library(kableExtra)

# --- Load config and data
config_data <- read_yaml(params$config_file)
samples_cfg  <- config_data$samples

read_mosdepth_summary <- function(s) {
    read.table(s$path, sep = "\t", header = TRUE) %>%
        filter(!grepl(chrom, pattern = "_")) %>%
        filter(!(chrom %in% c("chrM", "chrEBV", "total"))) %>%
        mutate(sample = s$id)
}

mosdepth <- lapply(samples_cfg, read_mosdepth_summary) %>% bind_rows()

chrom_order <- c(paste0("chr", 1:22), "chrX", "chrY")
mosdepth <- mosdepth %>%
    filter(chrom %in% chrom_order) %>%
    mutate(chrom = factor(chrom, levels = chrom_order))

# --- Compute per-sample summary metrics
summary_metrics <- mosdepth %>%
    group_by(sample) %>%
    summarise(
        total_bases_gb    = round(sum(bases) / 1e9, 2),
        median_coverage   = round(median(mean[!chrom %in% c("chrX", "chrY")]), 1),
        chrX_coverage     = round(mean[chrom == "chrX"], 1),
        chrY_coverage     = round(mean[chrom == "chrY"], 1),
        .groups = "drop"
    ) %>%
    mutate(
        sex_ratio    = chrX_coverage / median_coverage,
        inferred_sex = if_else(sex_ratio >= 0.7, "Female (XX)", "Male (XY)")
    ) %>%
    select(-sex_ratio)
```

---

## Overview

This report summarises the sequencing coverage quality control (QC) for the samples listed below.
It is generated automatically from [mosdepth](https://github.com/brentp/mosdepth) summary files
and is intended to give a quick overview of data quality before downstream analysis.

**What to look for:**

- **Total aligned bases** should be consistent across samples in the same run (~90–120 Gb for 30× WGS).
- **Coverage distribution** across autosomes (chr1–22) should be uniform. Chromosomes that fall far
  outside the box are worth investigating.
- **Sex chromosome coverage** immediately indicates biological sex and can reveal sample mix-ups:
  - *Female (XX)*: chrX ≈ autosomal coverage, chrY ≈ 0×
  - *Male (XY)*: chrX ≈ 0.5× autosomal coverage, chrY ≈ 0.5× autosomal coverage

---

## Sample Information

```{r sample-table}
data.frame(
    `Sample ID` = sapply(samples_cfg, `[[`, "id"),
    `Mosdepth file` = sapply(samples_cfg, `[[`, "path"),
    `File exists` = sapply(sapply(samples_cfg, `[[`, "path"), file.exists),
    check.names = FALSE
) %>%
kbl(align = c("l", "l", "c")) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
              full_width = TRUE) %>%
column_spec(3, color = if_else(
    sapply(sapply(samples_cfg, `[[`, "path"), file.exists), "green", "red"))
```

---

## QC Plots

### Total Aligned Bases

The bar chart shows the total number of bases that successfully aligned to the reference genome
(GRCh38) across chromosomes 1–22, X, and Y. This is a proxy for the amount of usable sequencing
data per sample.

> **Expected range for 30× WGS:** ~90–120 Gb per sample.
> Large differences between samples may indicate unequal sequencing depth or failed libraries.

```{r plot-aligned}
total_aligned <- mosdepth %>%
    group_by(sample) %>%
    summarise(total_bases_gb = sum(bases) / 1e9, .groups = "drop") %>%
    arrange(desc(total_bases_gb))

plot_ly(
    total_aligned,
    x = ~reorder(sample, -total_bases_gb),
    y = ~round(total_bases_gb, 2),
    type = "bar",
    hovertemplate = "<b>%{x}</b><br>%{y:.2f} Gb<extra></extra>",
    marker = list(color = ~total_bases_gb, colorscale = "Blues", showscale = FALSE)
) %>%
layout(
    xaxis = list(title = "Sample", tickangle = -45),
    yaxis = list(title = "Aligned bases (Gb)"),
    showlegend = FALSE
)
```

---

### Coverage Distribution

The boxplot shows the distribution of mean per-chromosome coverage across all autosomes (chr1–22).
Each point in the box represents one chromosome. Red diamonds mark the chrX and chrY coverage
values, which are plotted separately because they carry additional biological information (see
[Per-Chromosome Coverage](#per-chromosome-coverage) below).

> **What to look for:**
>
> - Boxes at consistent heights across samples indicate uniform sequencing depth.
> - Outlier chromosomes labelled inside the box may indicate repeat-rich or hard-to-sequence regions.
> - chrX below the box in one sample but inside the box in another may indicate a sex discordance.

```{r plot-boxplot}
mosdepth_auto <- mosdepth %>%
    group_by(sample) %>%
    mutate(median_coverage = median(mean[!chrom %in% c("chrX", "chrY")]))

sex_chrs <- mosdepth_auto %>% filter(chrom %in% c("chrX", "chrY"))

plot_ly(
    mosdepth_auto,
    x = ~reorder(sample, -median_coverage),
    y = ~mean,
    color = ~sample,
    type = "box",
    text = ~as.character(chrom),
    hovertemplate = "<b>%{text}</b><br>Coverage: %{y:.1f}×<extra></extra>",
    boxpoints = "outliers",
    pointpos = 0,
    showlegend = FALSE
) %>%
add_trace(
    data = sex_chrs,
    x = ~sample,
    y = ~mean,
    type = "scatter",
    mode = "markers+text",
    text = ~as.character(chrom),
    textposition = "top right",
    marker = list(size = 11, symbol = "diamond", color = "crimson"),
    name = "sex chromosomes",
    hovertemplate = "<b>%{text}</b><br>Coverage: %{y:.1f}×<extra></extra>",
    showlegend = TRUE,
    inherit = FALSE
) %>%
layout(
    xaxis = list(title = "Sample", tickangle = -45),
    yaxis = list(title = "Mean coverage (×)"),
    legend = list(orientation = "h", y = -0.25)
)
```

---

### Per-Chromosome Coverage {#per-chromosome-coverage}

This plot shows the mean coverage for every chromosome individually. The blue-shaded region
highlights the sex chromosomes (chrX and chrY). Viewing each sample as a line makes it easy to
spot chromosomes with systematically different coverage — for example a deletion (lower coverage)
or a trisomy (higher coverage).

> **Sex chromosome interpretation:**
>
> | Pattern | chrX | chrY | Inferred sex |
> |---------|------|------|-------------|
> | Female | ≈ autosomal | ~0× | XX |
> | Male | ≈ 0.5× autosomal | ≈ 0.5× autosomal | XY |

```{r plot-per-chr}
p <- plot_ly()
for (s in unique(mosdepth$sample)) {
    d <- filter(mosdepth, sample == s)
    p <- add_trace(p,
        x = ~chrom, y = ~mean, data = d,
        name = s,
        type = "scatter", mode = "lines+markers",
        marker = list(size = 7),
        hovertemplate = paste0("<b>%{x}</b><br>Coverage: %{y:.1f}×<extra>", s, "</extra>"))
}

layout(p,
    xaxis = list(title = "Chromosome", tickangle = -45),
    yaxis = list(title = "Mean coverage (×)"),
    showlegend = TRUE,
    shapes = list(
        list(type = "rect", xref = "x", yref = "paper",
             x0 = 21.5, x1 = 23.5, y0 = 0, y1 = 1,
             fillcolor = "rgba(135, 206, 250, 0.20)",
             line = list(width = 0))),
    annotations = list(
        list(x = 22.5, xref = "x", y = 1, yref = "paper",
             text = "sex chromosomes", showarrow = FALSE,
             yanchor = "bottom",
             font = list(size = 11, color = "steelblue")))
)
```

---

## Summary Metrics

Key QC metrics for each sample. Inferred sex is determined from the ratio of chrX coverage to
median autosomal coverage (ratio ≥ 0.7 → Female).

```{r summary-table}
summary_metrics %>%
    rename(
        `Sample`                  = sample,
        `Total bases (Gb)`        = total_bases_gb,
        `Median autosomal cov (×)` = median_coverage,
        `chrX cov (×)`            = chrX_coverage,
        `chrY cov (×)`            = chrY_coverage,
        `Inferred sex`            = inferred_sex
    ) %>%
datatable(
    rownames = FALSE,
    options  = list(pageLength = 25, dom = "t"),
    class    = "stripe hover compact"
) %>%
formatStyle(
    "Inferred sex",
    backgroundColor = styleEqual(
        c("Female (XX)", "Male (XY)"),
        c("#fce4ec", "#e3f2fd")
    )
)
```

---

*Report generated with [mosdepth](https://github.com/brentp/mosdepth) +
[Snakemake](https://snakemake.readthedocs.io) +
[Quarto](https://quarto.org).*
